Software Development in Startup Companies: The Green-field Startup Model

### 2013-02-21 article/main.tex
MUN review and comments

%%extra packages ------ 
\usepackage{longtable}   
\usepackage{float} %allows the use of [H] to force positioning of figures
\usepackage{subfigure}
\usepackage{hyperref} %used in references now  --- it shoud be removed
\usepackage{amssymb}
\usepackage[neveradjust]{paralist} %compact lists

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=4in]{figures/completemethodology2}
\caption{Grounded theory study process overview}\label{fig:gt:completemethodology}
\end{figure*}
%MUN: I think you can increase the font size in the figure a bit. Try thereby to keep the 
%height of the figure, expanding however the width (use horizontal space since it is a two 
%column figure).
%Write "Data analysis" horizontally

@@

Following the \textit{grounded theory} principles we attempted to capture the most relevant aspects of software development from startup practitioners, letting a theory emerge from the interviews and adjusting the research hypotheses and questions as we proceeded through. During these interviews we collected data related to engineering activities undertaken by startups in the time frame between the idea conception and the first open beta release of the product.  Then, we proceeded to the analysis of the data, finding important relations among concepts with a formal approach.

As suggested by Coleman, in view of the different versions of \textit{grounded theory} which emerged in the last years, researchers should indicate which \textit{``implementation''} of the theory is being used \cite{Coleman2007}. In fact, after the methodology was initially introduced, its original authors  had some divergences regarding how the theory should be formed by analyzing the data (Strauss and Corbin on one side, Glaser on the other).  Glaser advocated for a more \textit{puristic} approach, where the theoretical categories should \textit{``naturally emerge from the data''}, whilst Strauss and Corbin formulated an updated version of the methodology, which leaves to the researchers an higher degree of freedom. The latter approach empowers researchers' \textit{``theoretical sensitivity''} \cite{Corbin1990}, and encourages them to outline the research problem beforehand. Since the knowledge obtained from parallel execution of the SMS \cite{SMS} and our direct experience with startups companies provided 
a good initial level of knowledge, in this study we use the Corbin and Strauss approach \cite{Strauss1998}.

In this research we addressed technical aspects related to software development in startups, exploring their operational dynamics. In view of the lack of agreement on an unique definition of the word \textit{startup}, we delimited our initial contextual boundaries to newly created innovative and single-product software companies, in the time-frame that goes from the idea conception to the release of the first product in highly scalable markets. Moreover, since most startups are web-oriented companies \cite{tc-webstartups-link, Allen2003}, in our case study, we inquired web companies with little or no operating history.

@@

The initial design of the case study is supported by the results of the \textit{systematic mapping study} (SMS) \cite{SMS} which contributed to define the initial questions for both questionnaires and semistructured interviews. The process depicted in Figure \ref{fig:gt:completemethodology} is evolutionary and affects the design at each new iteration.  In \textit{data collection} the empirical results are integrated and stored in a single \textit{case study database} and subsequently processed in \textit{data analysis} to form the theoretical categories.  At each iteration the new emergent theory is updated following a formal procedure (\textit{Theory generation}), and after verifying that the \textit{theoretical saturation}\footnote{The point at which executing more interviews wouldn't bring any additional value for constructing the theory.} of categories has been achieved, we finally proceeded to the \textit{theory validation} or performed another complete cycle.

@@

The detail process is thoroughly described in the following subsections which are structured according to the five macro phases depicted in bold in Figure \ref{fig:gt:completemethodology}.

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Design and execution}

This case study started based on a Systematic Mapping Study (SMS) of the literature regarding software development in startups \cite{SMS}. By means of this SMS we refined our research questions. The research questions have been narrowed down to outline the domain of our study, even though still broad enough to allow emerging new theories and to adjust the scope of the research \cite{Corbin1990}.

The sampling strategy took place in two distinct phases. At first we executed an initial \textit{convenience sampling} \cite{Dawson2009}, which led to the identification of eight companies. Then we included five additional startups during the theory formation process (\textit{theoretical sampling}) iteratively improving the sample according to the emerging theory.  Additional information about the companies are reported in Table \ref{t_interviews-stats}.

@@

As shown in column \textit{Company age}, most of the companies were founded within the last three years with only one exception (C10, which age was 3 years and 6 months). As can be observed by looking at the column \textit{Founding team}, the companies have been founded by an average number of  3 members, which in majority are developers, as can be seen by looking at the column \textit{Initial developers}. Moreover, the number of \textit{Current employees} shows how, to different degrees, companies expanded the initial teams, being able to bring the product to the market in less than 6 months from the idea conception\footnote{C5 building time is extended to 1 year mainly because the product has been developed part-time by one person only.} (see \textit{First product building time}).
Summarizing the above identified features, the sample of the case study is mainly composed by recently founded companies, working with web applications in different nations and market sectors, with very small development teams. Moreover, the growing team size and and the publicly available data suggest an healthy status of the businesses.

We executed the case study online, supported by tools for video conferencing recording each complete session. Each interview has been executed only once per company with the interviewee's role of CEO or CTO. A step-by-step workflow have been followed, which allowed us to perform the interviews, collect additional material, prepare the customized follow-up questionnaire and iteratively adjust the \textit{interview package} artifacts\footnote{The \textit{interview package} and {follow-up questionnaire} has been released under MIT License \cite{MITLicense} and it is available for download on Github \cite{GitHubInterviewPackage}. We encourage anyone who has interests in pushing this work forward, to fork the repository and contribute to it. If used in a research context, please inform us in order to be able to track where and how the \textit{interview package} has been used.}.

The follow-up questionnaires were designed to capture additional data, gather missing information and confirm interview results with triangulation. Questionnaires have been tailored to each startup, partially taking advantage of the repertory grid principles \cite{Edwards2009}. To obtain better chances of quickly obtaining responses, questionnaires have been sent to respondents immediately after transcribing the interview results. Two weeks after the conclusion of the last interview, we closed our follow-up questionnaires and collected the data.

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Data collection}
The approach undertaken to collect the data, known as \textit{data triangulation}, integrates multiple data sources converging on the same phenomenon.
After transcribing the interview, we extracted conceptual relations, and then we integrated them with the questionnaire's results. A well structured case-study database allowed us to easily retrieve and seek for information, assembling the evidence from the case study reports, as described also in \cite{Yin1994}. The database has been stored and constructed using the qualitative data analysis software package AtlasTI\footnote{Available online at \url{http://www.atlasti.com/} .} - one of the most suitable tools for GT \cite{Coleman2007}.

Entering the field, we overlapped interviews with questionnaire results to adjust the data collection and take advantage of emergent themes and reveal possible inconsistencies. The data was analyzed simultaneously and with flexibility in mind in such a way that adjustments were made according to the emerging findings.
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Data analysis}
Before starting the analysis, a \textit{data ordering} procedure was necessary since interviews were spread across a multitude of topics. Therefore transcripts have been structured in thematic areas accordingly to different \textit{topic cards} used during the interviews. We proceeded horizontally between same thematic areas of transcripts to be able to identify a better number of similar concepts, rather than going through an entire transcript at one time.

Once the data were ordered, we executed the process of \textit{coding} interviews, following the steps listed below:

@@

\item Labels were assigned to raw data, and a first low-level conceptualization was carried out using both \textit{in-vivo} and \textit{open coding} \cite{ColinRobson2009}.
\item Concepts were grouped together into theoretical categories and subcategories. By means of \textit{axial coding} we first described the different relations between subcategories, and then relations between subcategories and categories.
\item Categories were refined several times in the attempt to create different level of abstractions and adjusting concepts, aided by a simple knowledge management tool.
\item Consistency among categories were validated by exploring and analyzing links  among subcategories by means of the \textit{selective coding}.
\item The \textit{core category} - the one with the greatest explanatory power -  was identified by analyzing the causal relations between high-level categories.
\end{compactitem}

During data extraction we used the technique of \textit{in-vivo coding} since \textit{``direct quoting from the transcript give more expressive power to the data''} \cite{ColinRobson2009} combined with the more descriptive procedure of \textit{open coding}. Following the example of other grounded theories, developed in related areas such as Information Systems \cite{Orlikowski1993} and Software Process Improvement \cite{Coleman2006},  we performed the high-level conceptualization during creation of categories, in the process of refining axial and selective coding.

As we were iterating through the interviews, we analyzed new data accordingly, by updating \textit{codes} and \textit{categories} on necessity, and taking notes in the form of  memos to adjust the new \textit{emerging theory}. To improve the speed of the coding process, we analyzed all the transcripts one part at-the-time: transcripts followed the semi-structured interview format, and this approach enhanced the chances of finding similar codes across different transcripts, allowing us to quickly iterate and update categories.

Given the large number of codes, categories, properties, propositions and related questions that evolved from the analytical process \cite{Corbin1990}, an important activity that helped the coding process was \textit{memoing}. It constituted an important component involved in the formulation and revision of theory during the research process. For this purpose we made use of three different types of memos: code memos, theoretical memos and operational memos. The first type of memo is related to the conceptual labeling of open codes, whilst the second type concerns axial and selective coding, and thus focusing on paradigm features. Finally, operational memos contain directions relating to the evolving \textit{emerging theory}.

After the coding process the first representation of the \textit{experience map} identified in GT is constructed by means of a \textit{theoretical framework}. The theoretical framework is presented in the form of a network of categories and subcategories identified during the coding process. Precisely, categories and subcategories are linked together according to cause-effect relation \cite{Corbin1990}. During the formation of the theoretical framework the researchers operated by means of a  bottom-up approach. From empirical data and coding process, the framework was developed into two different levels: a detailed level representing the network of subcategories (identified mainly by \textit{axial coding} process), and high-level representing the network of main categories (identified mainly by \textit{selective coding} process).

@@

As mentioned above, emergent theories have been tested integrating the sample with additional companies selected following the principle of \textit{theoretical sampling} \cite{Yin1994}.

The process of \textit{theory generation} took place at each iteration together with interview execution, where we systematically analyzed the \textit{theoretical framework} by means of the \textit{paradigm model} introduced by Corbin \cite{Corbin1990}.

The paradigm model is composed by:

\begin{compactitem}
\item \textit{Causal conditions}, the events which lead to the occurrence of the \textit{phenomenon}, that is our core category.
\item \textit{Context}, set of conditions in which the phenomenon can be extrapolated;
\item \textit{Intervening conditions}, the broader set of condition that the phenomenon can be generalized.
\item \textit{Action/interaction strategies}, the actions and responses that occur as the result of the phenomenon.
\item \textit{Consequences}, specification of the outcomes, both intended and unintended of the actions and interaction strategies.
\end{compactitem}

Within the limits of the critical bounding assumptions, the role of the generated theory is to explain, predict and understand the underling phenomenon.

@@

By analyzing the interview transcripts with the GT process we extracted the information necessary to form the theoretical model. The initial coding process (open and in-vivo) provided 630 unique codes \footnote{Available at \url{https://github.com/adv0r/BTH-Interview-Package}.}, grounded to 1295 citations in the transcripts, divided according to the thematic areas.

The thematic areas were divided according to the development phases designed for the interview process. The complete list of  raw codes is presented in Appendix and constitutes the empirical foundation of the theoretical categories, which has been used to form the theoretical framework (Green-field startup model).

In this section we present the GSM model to capture the underlying phenomenon of software development in early-stage startups. The GSM model is based on a systematic procedure which is grounded into empirical concepts obtained by conduction of the case-study. Starting from 630 raw \textit{codes} extrapolated from the interview transcripts, the framework is formed by 128 sub-categories, clustered in 35 \textit{groups}, and finally in 7 categories at the highest level of abstraction.

By the means of a conceptual framework we want to provide explanations of the development strategies and engineering activities undertaken by startups. The knowledge, which we attempted to pack into a conceptual model, can be used by researchers and practitioners to improve their understanding of the phenomenon.
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Framework overview}
\label{res:gsm:frmov}
Main concepts representing underlying phenomena have been grouped together to form high-level categories. Figure \ref{fig:gsm} shows the network of causal relationships (represented by arrows) between categories (represented by blocks). Each node is linked to subsequent nodes which are called successors, and preceding nodes which are called predecessors. The relations between them are denoted by the direction of the interconnecting arrows. The network is read from left to right.

@@

In the explanation of the GSM model we emphasize labels of  theoretical categories using the \textit{italic} font style. For the main categories shown in Figure \ref{fig:gsm}, when necessary we also make use of identifiers (i.e. CATx) to facilitate the reading of this section.
The network is centered around the core category, \textit{speed-up development}, which is the most interconnected node in the framework reflecting the fact that \textit{``it is the one (category) with the greatest explanatory power''}  \cite{Strauss1998}. The meaning of the labels on the three arrows that reach the core category (\textit{efficiency}, \textit{effectiveness} and \textit{performance}) is discussed in detail in section \ref{valt}.

A contextual condition which characterizes, to some extent, each and every startup, is the \textit{severe lack of resources}. In fact, the capabilities of an early-stage startups' to support its development activities are constrained by an extremely limited access to human, time and intellectual resources, eventually leading to a general absence of structures and processes.

The \textit{severe lack of resources} forces the company to focus on implementing an essential set of functionalities and it is one of the main reasons why the \textit{product quality has low priority} respect to other more urgent needs\footnote{There are some exceptions where the quality aspects actually matter and such cases will be discussed later in \ref{valt}.}. At the same time, to be able to deal with such constraints, startups must depend on a small group of capable and motivated individuals.

@@

The initial employees are the ingredients which enable high levels of performance in software development. To support a fast-paced production environment, engineers are required to be highly committed, co-located, multi-role, and self-organized. In other words \textit{team is the catalyst of development}.
With an essential and flexible workflow, which relies on tacit knowledge instead of formal documentation, startups are able to achieve very short time-to-market. However, each line of code, written without following structures and processes, contributes to grow the \textit{accumulated technical debt}, which is further increased by having almost non-existing specifications, a minimal project management  and a severe lack of automated tests.

@@

In the subsequent subsections we are going to explain in detail all the categories presented in Figure \ref{fig:gsm}, to conclude in subsection \ref{res:gsm:th} with the final theory. During the following explanations we are going to use identifiers of the companies presented in Table \ref{t_interviews-stats} (i.e. Cx) to highlight statements recorded from the interviewees.

@@

In startups \textit{developers have big responsibilities}. In fact, the \textit{limited human resources}, discussed in CAT7, causes the team-members to be active in every aspect of the development process, from the definition of functionalities to the final deployment.

Early engineers in the founding team of a startups are typically \textit{multi-role and full-stack engineers}: they are full-stack for being able to tackle different problems at different level of the technology stack (\textit{generalists developers instead of specialists}) and multi-role since usually \textit{engineers are responsible for marketing/sales/development}. As remarked by  C11: \textit{``Instead of hiring gurus in one technology, startups should hire young developers, generalists, that know how to quickly learn new technologies, and quickly move among a huge variety of tasks.''}.

Moreover, having a \textit{very small and co-located development team} enables members to operate with high coordination, control and communication, which heavily rely on \textit{tacit knowledge},  replacing most of the documentation with informal discussions. Practitioners reported that keeping the development team small helps startups in being fast and flexible, as remarked by C8: \textit{``If you have more than 10 people, it is absolutely impossible to be fast''}. Then, also \textit{previous working experience} and \textit{knowing each other before starting the company} enforce the efficiency of activities by \textit{limiting th need of formalities between team-members}.

In every software company, \textit{skilled developers are essential for high speed} in development. Especially in startups, the ``hacking culture'' and a tendency to the ``just-do-it'' approach allow the team to quickly move from the formulation of a feature idea to its implementation. In this regard, C1 comments: \textit{``we had a hacker culture/environment, people hacking stuff without formally analyzing it, but breaking it down and finding a way around. Ticket on trello and go.''}

A \textit{limited access to expertise} forces the team to rely mainly on their personal abilities, even though asking opinions to mentors is reported to be a viable practice to maintain feasible objectives. Furthermore \textit{teams work under constant pressure}, \textit{``up to 16 hours per day''}, mainly constrained by a tight \textit{time shortage}.

@@

Since \textit{flexibility and reactiveness are the main priorities}, the most suitable class of software development approaches are clearly highly evolutionary in nature. In fact, as \textit{uncertain conditions make long-term planning not viable}, startups cannot base their work on assumptions without rapidly validating them, releasing the product to market. The uncertainty is first of all in the team composition. In fact, since the teams are typically very small and the project knowledge is not documented, even a little change in their composition (e.g. a developer falls in ill) can have a very high impact on the overall product development. Furthermore, startups operate in a continuous evolving environment in terms of competitors and targeted market sectors. Then, to obtain a competitive advantage in the market, startups typically make use of cutting-edge solutions, characterised by an evolution that cannot be foreseen in the long run. However, a special role in daily decisions is covered by user feedback 
and requests, which represent the main drivers to define the product features to provide in the short-term.

To obtain fast user responses and quickly \textit{validate the product}, startups \textit{build a functioning prototype and iterate on it} over time. In fact, paraphrasing C4, \textit{``[\ldots] you should start with something that is really rough and then polish it, fix it and iterate. We were under constant pressure. The aim was to understand as soon as possible the product market/fit iterating quickly, adjusting the product and releasing fast.''}

The company focus on building a small set of functionalities to include in the first version, and \textit{progressively rolls-out to larger number of people} with very \textit{small iterations} (confirmed again by C4: \textit{``we deploy from 5 to 20 times a day''}).

The scope of this evolutionary approach is to avoid wasting time in ``over-engineering the system'' and building complex functionalities that have not been tested on real users. By releasing a very small number of good-enough functionalities (see CAT3) the startup can verify the suitability of the features and understand how to adjust the direction of product development towards actual users' needs. The first version of the product is typically a prototype which contains the basic rough functionalities developed with the least possible effort that can validate critical features which can undermine the startup's survival in the short term (recalling  Ries' definition of \textit{Minimum Viable Product} (MVP) \cite{Ries2011}).

>>>>

%%extra packages ------ 
\usepackage{longtable}   
\usepackage{float} %allows the use of [H] to force positioning of figures
\usepackage{subfigure}
\usepackage{hyperref} %used in references now  --- it shoud be removed
\usepackage{amssymb}
\usepackage[neveradjust]{paralist} %compact lists
\usepackage{amsthm}

\newtheorem*{theory}{Theory}

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=4in]{figures/completemethodology2}
\caption{Grounded theory study process overview}\label{fig:gt:completemethodology}
\end{figure*}
%MUN: I think you can increase the font size in the figure a bit. Try thereby to keep the 
%height of the figure, expanding however the width (use horizontal space since it is a two 
%column figure). Change also in figure:
% - Write "Data Analysis" horizontally
% - Change names of macro-phases to upper-case, i.e. Design and Execution, Data Collection, Data Analysis, etc.
% - Add section number to macro-phase name, i.e. 3.1 Design and Execution, 3.2 Data Collection
% - Change "Discussion and Validation" into "Theory Validation" as you name it in the text.
% Also, add a section (3.5) that describes shortly what you did in "Theory Validation"

@@

Following the GT principles we attempted to capture the most relevant aspects of software development from startup practitioners, letting a theory emerge from the interviews and adjusting the research hypotheses and questions as we proceeded through. During these interviews we collected data related to engineering activities undertaken by startups in the time frame between the idea conception and the first open beta release of the product.  Then, we proceeded to the analysis of the data, finding important relations among concepts with a formal approach.

As suggested by Coleman, in view of the different versions of GT which emerged in the last years, researchers should indicate which \textit{``implementation''} of the theory is being used \cite{Coleman2007}. After the methodology was initially introduced, its original authors  had divergences regarding how the theory should be formed by analyzing the data (Strauss and Corbin on one side, Glaser on the other).  Glaser advocated for a more \textit{puristic} approach, where the theoretical categories should \textit{``naturally emerge from the data''}, whilst Strauss and Corbin formulated an updated version of the methodology, which leaves to the researchers an higher degree of freedom. The latter approach empowers researchers' \textit{``theoretical sensitivity''} \cite{Corbin1990}, and encourages them to outline the research problem beforehand. Since the knowledge obtained from parallel execution of the SMS \cite{SMS} and our direct experience with startups companies provided 
a good initial level of knowledge, in this study we use the Corbin and Strauss approach \cite{Strauss1998}.

In this research we address technical aspects related to software development in startups, exploring their operational dynamics. In view of the lack of agreement on an unique definition of the word \textit{startup}, we delimited our initial contextual boundaries to newly created innovative and single-product software companies, in the time-frame that goes from the idea conception to the release of the first product in highly scalable markets. Moreover, since most startups are web-oriented companies \cite{tc-webstartups-link, Allen2003}, in our case study, we inquired web companies with little or no operating history.

@@

The initial design of the case study is supported by the results of the \textit{Systematic Mapping Study} (SMS) \cite{SMS} which contributed to define the initial questions for both questionnaires and semi-structured interviews. The process depicted in Figure \ref{fig:gt:completemethodology} is evolutionary and affects the design at each new iteration. In \textit{Data Collection} the empirical results are integrated and stored in a single case study database and subsequently processed in \textit{Data Analysis} to form the theoretical categories.  At each iteration the new emergent theory is updated following a formal procedure (\textit{Theory Generation}), and after verifying that \textit{Theoretical Saturation}\footnote{The point at which executing more interviews wouldn't bring any additional value for constructing the theory.} of categories has been achieved, we finally proceeded to \textit{Theory Validation}.

@@

The process details are described in the following subsections, structured according to the five macro phases depicted in bold in Figure \ref{fig:gt:completemethodology}.

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Design and Execution}
\label{desex}

This case study started based on a SMS of the literature regarding software development in startups \cite{SMS}. By means of this SMS we refined our research questions. The research questions have been narrowed down to outline the domain of our study, even though still broad enough to allow emerging new theories and to adjust the scope of the research \cite{Corbin1990}.

The sampling of candidate companies took place in two distinct phases. At first we executed an initial convenience sampling \cite{Dawson2009}, which led to the identification of eight companies. Then we included five additional startups during the theory formation process (theoretical sampling) iteratively improving the sample according to the emerging theory. The characteristics of the sampled companies are reported in Table \ref{t_interviews-stats}.

@@

All companies, except C10, were founded within the last three years (2009-2012), %MUN: I've shortened the text and
%added absolute numbers. Is 2009-2012 correct?
by an average of 3 starting members, which were in majority developers. Moreover, the number of current employees shows how, to different degrees, companies expanded the initial teams. All companies, except C5, brought their first product to market in less than 6 months from the idea conception. The products consist of web-applications, launched in different nations and market sectors. The growing team size and the publicly available data suggest an healthy status of the businesses.

We executed the case study online, supported by tools for video conferencing, recording each complete session. The interview subjects were the companies' CEOs or CTOs. We followed a step-by-step workflow, consisting of the actual interview, collection of additional material, preparation of the customized follow-up questionnaire and the iterative adjustment of the interview package artifacts\footnote{The interview package and follow-up questionnaire has been released under MIT License \cite{MITLicense} and it is available for download on Github \cite{GitHubInterviewPackage}. We encourage anyone interested in pushing this work forward, to fork the repository and contribute to it. If used in a research context, please inform us in order to be able to track where and how the interview package has been used.}.

The follow-up questionnaires were designed to capture additional data, gather missing information and confirm interview results by triangulation. The questionnaires have been tailored to each startup, partially taking advantage of the repertory grid principles \cite{Edwards2009}. To obtain better chances of quickly obtaining responses, questionnaires have been sent to respondents immediately after transcribing the interview results. Two weeks after the conclusion of the last interview, we closed our follow-up questionnaires and collected the data. %MUN: I don't understand this: you
% collected the data from questionnaires after ALL interviews have been finished, but on the other hand you 
% said the overall process was iterative (data analysis and triangulation let themes emerge which in turn were used
% in subsequent interviews).

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Data collection}
The data collection was designed to allow triangulation, which integrates multiple data sources converging on the same phenomenon. After transcribing the interview, we extracted conceptual relations, integrating them with the questionnaire's results. A well structured case-study database allowed us to easily retrieve and seek for information, assembling the evidence from the case study reports, as described also in \cite{Yin1994}. The database has been stored and constructed using the qualitative data analysis software package AtlasTI\footnote{Available online at \url{http://www.atlasti.com/} .} - one of the most suitable tools for GT \cite{Coleman2007}. 
We overlapped interviews with questionnaire results to adjust the data collection and take advantage of emergent themes and reveal possible inconsistencies. The data was analyzed simultaneously and with flexibility in mind in such a way that adjustments were made according to the emerging findings.
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Data analysis}
Before starting the analysis, a data ordering procedure was necessary since interviews were spread across a multitude of topics. Therefore transcripts have been structured in thematic areas accordingly to different topic cards used during the interviews. We proceeded horizontally between same thematic areas of transcripts to be able to identify a better number of similar concepts, rather than going through an entire transcript at one time.

Once the data were ordered, we executed the process of coding interviews, following the steps listed below:

@@

\item Labels were assigned to raw data, and a first low-level conceptualization was carried out using both in-vivo and open coding \cite{ColinRobson2009}.
\item Concepts were grouped together into theoretical categories and subcategories. By means of axial coding we first described the different relations between subcategories, and then relations between subcategories and categories.
\item Categories were refined several times in the attempt to create different levels of abstraction and adjusting concepts, aided by a simple knowledge management tool.
\item Consistency among categories was validated by exploring and analyzing links among subcategories by means of selective coding.
\item The core category - the one with the greatest explanatory power -  was identified by analyzing the causal relations between high-level categories.
\end{compactitem}

During data extraction we used the technique of in-vivo coding since \textit{``direct quoting from the transcript gives more expressive power to the data''} \cite{ColinRobson2009} combined with the more descriptive procedure of open coding. Following the example of other grounded theories, developed in related areas such as Information Systems \cite{Orlikowski1993} and Software Process Improvement \cite{Coleman2006}, we performed the high-level conceptualization during creation of categories, in the process of refining axial and selective coding.

As we were iterating through the interviews, we analyzed new data accordingly, by updating codes and categories on necessity, and taking notes in the form of memos to adjust the new emerging theory. To improve the speed of the coding process, we analyzed all the transcripts one part at-the-time: transcripts followed the semi-structured interview format, and this approach enhanced the chances of finding similar codes across different transcripts, allowing us to quickly iterate and update categories.

Given the large number of codes, categories, properties, propositions and related questions that evolved from the analytical process \cite{Corbin1990}, an important activity that helped the coding process was memoing. It constituted an important component involved in the formulation and revision of theory during the research process. For this purpose we made use of three different types of memos: code memos, theoretical memos and operational memos. The first type of memo is related to the conceptual labeling of open codes, whilst the second type concerns axial and selective coding, and thus focusing on paradigm features. Finally, operational memos contain directions relating to the evolving emerging theory.

After the coding process the first representation of the experience map identified in GT is constructed by means of a theoretical framework. The theoretical framework is presented in the form of a network of categories and subcategories that are linked together according to cause-effect relationships \cite{Corbin1990}. During the formation of the theoretical framework the researchers operated by means of a  bottom-up approach. From the empirical data and coding process, the framework developed into two different levels: a detailed level representing the network of subcategories (identified mainly by the axial coding process), and a high-level representation of the main categories network (identified mainly by the selective coding process).

@@

As mentioned in Section \ref{desex}, emergent theories have been tested by integrating additional companies into the sample, selected following the principle of theoretical sampling \cite{Yin1994}.

The process of theory generation took place at each iteration together with interview execution, where we systematically analyzed the theoretical framework by means of the paradigm model introduced by Corbin \cite{Corbin1990}. The paradigm model is composed by:

\begin{compactitem}
\item \textit{Causal conditions}: the events which lead to the occurrence of the phenomenon, that is our core category
\item \textit{Context}: set of conditions in which the phenomenon can be extrapolated
\item \textit{Intervening conditions}: the broader set of conditions from that the phenomenon can be generalized
\item \textit{Action/interaction strategies}: the actions and responses that occur as the result of the phenomenon
\item \textit{Consequences}: specification of the outcomes, both intended and unintended of the actions and interaction strategies
\end{compactitem}

Within the limits of the critical bounding assumptions, the role of the generated theory is to explain, predict and understand the underlying phenomenon.

\subsection{Theory Validation}
%MUN: put here what I wrote in the comment in Section "Theory validation" (i.e. the section following the GSM)

@@

By analyzing the interview transcripts with the GT process we extracted the information necessary to form the theoretical model. The initial coding process (open and in-vivo) provided 630 unique codes \footnote{Available at \url{https://github.com/adv0r/BTH-Interview-Package}.}, grounded to 1295 citations in the transcripts, divided according to the thematic areas. The thematic areas were divided according to the development phases designed for the interview process. The raw codes constitute the empirical foundation of the theoretical categories, which has been used to form the Green-field startup model (GSM).

The GSM captures the underlying phenomenon of software development in early-stage startups. The framework is formed by 128 sub-categories, clustered in 35 groups, and finally in 7 categories at the highest level of abstraction. By the means of the GSM we provide explanations of the development strategies and engineering activities undertaken by startups.
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Framework overview}
\label{res:gsm:frmov}
The main concepts representing the underlying phenomenon have been grouped together to form high-level categories. Figure \ref{fig:gsm} shows the network of causal relationships (represented by arrows) between categories (represented by blocks). Each node is linked to subsequent nodes which are called successors, and preceding nodes, called predecessors. The relations between them are denoted by the direction of the interconnecting arrows. The network is read from left to right.

@@

In forthcoming the explanation of the GSM we emphasize labels of theoretical categories using the \textit{italic} font style. For the main categories shown in Figure \ref{fig:gsm}, when necessary we also make use of identifiers (i.e. CATx) to facilitate the reading of this section.
The network is centered around the core category, \textit{speed-up development}, which is the most interconnected node in the framework reflecting the fact that \textit{``it is the one (category) with the greatest explanatory power''}  \cite{Strauss1998}. The meaning of the labels on the three arrows that reach the core category (efficiency, effectiveness and performance) is discussed in detail in section \ref{valt}. %MUN: this refers to the validity threats section. Do you mean rather the "Theory Validation" section?

A contextual condition which characterizes, to some extent, each and every startup, is the \textit{severe lack of resources}. In fact, the capabilities of an early-stage startup to support its development activities are constrained by an extremely limited access to human, time and intellectual resources, eventually leading to a general absence of structures and processes.

The \textit{severe lack of resources} forces the company to focus on implementing an essential set of functionalities and it is one of the main reasons why the \textit{product quality has low priority} respect to other more urgent needs\footnote{There are some exceptions where the quality aspects actually matter and such cases will be discussed in section \ref{valt} %MUN: this refers to the validity threats section. Do you mean rather the "Theory Validation" section?
.}. At the same time, to be able to deal with such constraints, startups must depend on a small group of capable and motivated individuals.

@@

The initial employees are the ingredients which enable high levels of performance in software development. To support a fast-paced production environment, engineers are required to be highly committed, co-located, multi-role, and self-organized. In other words \textit{team is the catalyst of development}.
With an essential and flexible workflow, which relies on tacit knowledge instead of formal documentation, startups are able to achieve very short time-to-market cycles. However, each line of code, written without following structures and processes, contributes to grow the \textit{accumulated technical debt}, which is further increased by having almost non-existing specifications, a minimal project management and a lack of automated tests.

@@

In the following subsections we are going to explain in detail all the categories presented in Figure \ref{fig:gsm}, and conclude in subsection \ref{res:gsm:th} with the final theory. In the explanations we use identifiers of the companies presented in Table \ref{t_interviews-stats} (i.e. Cx) to highlight statements recorded from the interviewees.

@@

In startups \textit{developers have big responsibilities}. In fact, the \textit{limited human resources}, discussed in CAT7, cause the team-members to be active in every aspect of the development process, from the definition of functionalities to the final deployment.

Early engineers in the founding team of a startups are typically \textit{multi-role and full-stack engineers}: they are full-stack for being able to tackle different problems at different levels of the technology stack (\textit{generalists developers instead of specialists}) and multi-role since usually \textit{engineers are responsible for marketing/sales/development}. As remarked by  C11: \textit{``Instead of hiring gurus in one technology, startups should hire young developers, generalists, that know how to quickly learn new technologies, and quickly move among a huge variety of tasks.''}

Moreover, having a \textit{very small and co-located development team} enables members to operate with high coordination, control and communication, which heavily rely on \textit{tacit knowledge},  replacing most of the documentation with informal discussions. Practitioners reported that keeping the development team small helps startups in being fast and flexible, as remarked by C8: \textit{``If you have more than 10 people, it is absolutely impossible to be fast''}. Then, also \textit{previous working experience} and \textit{knowing each other before starting the company} enforce the efficiency of activities by \textit{limiting the need of formalities between team-members}.

In every software company, \textit{skilled developers are essential for high speed} in development. Especially in startups, the ``hacking culture'' and a tendency to the ``just-do-it'' approach allow the team to quickly move from the formulation of a feature idea to its implementation. In this regard, C1 comments: \textit{``We had a hacker culture/environment, people hacking stuff without formally analyzing it, but breaking it down and finding a way around. Ticket on trello and go.''}

A \textit{limited access to expertise} forces the team to rely mainly on their personal abilities, even though asking opinions to mentors is reported to be a viable practice to maintain feasible objectives. Furthermore \textit{teams work under constant pressure}, \textit{``up to 16 hours per day''}, %MUN: add Cx to this statement since it is a citation?
mainly constrained by a tight \textit{time shortage}.

@@

Since \textit{flexibility and reactiveness are the main priorities}, the most suitable class of software development approaches are clearly highly evolutionary in nature. In fact, as \textit{uncertain conditions make long-term planning not viable}, startups cannot base their work on assumptions without rapidly validating them, releasing the product to market. The uncertainty is first of all in the team composition. Since the teams are typically very small and the project knowledge is not documented, even a little change in their composition (e.g. a developer falls ill) can have a very high impact on the overall product development. Furthermore, startups operate in a continuous evolving environment in terms of competitors and targeted market sectors. Then, to obtain a competitive advantage in the market, startups typically make use of cutting-edge solutions, characterised by an evolution that cannot be foreseen in the long run. However, a special role in daily decisions is covered by user feedback 
and requests, which represent the main drivers for defining the product features in the short-term.

To obtain fast user responses and quickly \textit{validate the product}, startups \textit{build a functioning prototype and iterate on it} over time. Paraphrasing C4, \textit{``[\ldots] you should start with something that is really rough and then polish it, fix it and iterate. We were under constant pressure. The aim was to understand as soon as possible the product market/fit iterating quickly, adjusting the product and releasing fast.''}

The companies focus on building a small set of functionalities to include in the first version, and \textit{progressively roll-out to a larger number of people} with very \textit{small iterations} (confirmed again by C4: \textit{``we deploy from 5 to 20 times a day''}).

The objective of this evolutionary approach is to avoid wasting time in ``over-engineering the system'' and building complex functionalities that have not been tested on real users. By releasing a very small number of good-enough functionalities (see CAT3) the startup can verify the suitability of the features and understand how to adjust the direction of product development towards actual users' needs. The first version of the product is typically a prototype which contains the basic functionalities developed with the least possible effort that can validate critical features, enabling the startup's survival in the short term (recalling  Ries' definition of \textit{Minimum Viable Product} (MVP) \cite{Ries2011}).

Supported by \textit{direct contact and observation of users}, \textit{automated feedback collection} and \textit{analysis of product metrics}, startups attempt to \textit{find what is valuable for customers} with activities in agreement with Blank's \textit{Customer validation} stage \cite{Blank2005}.
